---
title: "GOA SST MHW by depth"
author: "Matt Callahan"
date: '2022-12-15'
output: word_document
---

## GOA and AI SST and MHW by depth bin
The purpose of this analysis is to compare climatology and heatwave status between depth strata in the Gulf of Alaska and Aleutian Islands. 
We first query a depth specific sst time series for each ESR subregion.
Depth strata are 10-200m, 201-1000m, and >1000m deep.
We then plot SST for each depth strata in each subregion.
Finally we calculate heatwave status for each depth strata/subregion

## SST download
Here we calculate daily average SST for GOA and AI ESR subregions. 

**connect to AKFIN**
```{r, eval=FALSE, include=FALSE}
library(odbc)
library(tidyverse)
library(getPass)

#connect
con <- dbConnect(odbc::odbc(), "akfin", UID=getPass(msg="USER NAME"), PWD=getPass())
```

**query SST**
```{r, eval=FALSE, include=FALSE}
#extend timeout

#-10 to -200
sst_10_200<- dbFetch(dbSendQuery(con,
                           paste0("select b.ecosystem_sub, a.read_date, avg(a.temp) mean_sst from afsc.erddap_crw_sst a
left join afsc.erddap_crw_sst_spatial_lookup b on a.crw_id = b.id
where b.depth between -200 and -10
and b.ecosystem in ('Gulf of Alaska' , 'Aleutian Islands')
group by b.ecosystem_sub, a.read_date
order by b.ecosystem_sub, a.read_date")))

sst_10_200<-sst_10_200 %>% mutate(depth_strata="-10 to -200")

saveRDS(sst_10_200, "data/sst_10_200.RDS")

#-201 to -1000
sst_201_1000<- dbFetch(dbSendQuery(con,
                           paste0("select b.ecosystem_sub, a.read_date, avg(a.temp) mean_sst from afsc.erddap_crw_sst a
left join afsc.erddap_crw_sst_spatial_lookup b on a.crw_id = b.id
where b.depth between -1000 and -201
and b.ecosystem in ('Gulf of Alaska' , 'Aleutian Islands')
group by b.ecosystem_sub, a.read_date
order by b.ecosystem_sub, a.read_date")))

sst_201_1000<-sst_201_1000 %>% mutate(depth_strata="-201 to -1000")

saveRDS(sst_201_1000, "data/sst_201_1000.RDS")

# >1000
sst_1001<- dbFetch(dbSendQuery(con,
                           paste0("select b.ecosystem_sub, a.read_date, avg(a.temp) mean_sst from afsc.erddap_crw_sst a
left join afsc.erddap_crw_sst_spatial_lookup b on a.crw_id = b.id
where b.depth < -1000
and b.ecosystem in ('Gulf of Alaska' , 'Aleutian Islands')
group by b.ecosystem_sub, a.read_date")))

sst_1001<-sst_1001 %>% mutate(depth_strata="< -1000")

saveRDS(sst_1001, "data/sst_1001.RDS")

#spatial lookup table for plotting
dbFetch(dbSendQuery(con, paste0("select* from afsc.erddap_crw_sst_spatial_lookup"))) %>%
  saveRDS("data/crw_spatial_lookup.RDS")
```

## Generate map showing regions
This map is pretty rough. The narrowness of the shelf makes the boundaries difficult to distinguish.

```{r, warning=FALSE, message=FALSE}
library(lubridate)
library(heatwaveR)
library(sf)
library(akmarineareas2)
library(tidyverse)

#original version that included aleutians
# lkp<-readRDS("data/crw_spatial_lookup.RDS") %>%
#   rename_with(tolower)%>%
#   filter(depth <= -10 & ecosystem %in% c("Gulf of Alaska", "Aleutian Islands")) %>%
#   mutate(depth_strata=ifelse(depth>=-200, "-10 to -200", ifelse(depth>= -1000, "-201 to -1000", "< -1000")),
#          lon360=ifelse(longitude>0, longitude, longitude+360))%>%
#   st_as_sf(coords = c('longitude', 'latitude'), crs = 4326, agr = 'constant') %>%
#   st_transform(crs=3338)

#New version with just GOA
lkp<-readRDS("data/crw_spatial_lookup.RDS") %>%
  rename_with(tolower)%>%
  filter(depth <= -10 & ecosystem =="Gulf of Alaska") %>%
  mutate(depth_strata=ifelse(depth>=-200, "-10 to -200", ifelse(depth>= -1000, "-201 to -1000", "< -1000")),
         lon360=ifelse(longitude>0, longitude, longitude+360))%>%
  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326, agr = 'constant') %>%
  st_transform(crs=3338)

#map
ggplot()+
 # geom_sf(data=ak_dd%>%st_shift_longitude())+
  #geom_point(data=lkp, aes(x=lon360, y= latitude, color=depth_strata, fill= depth_strata))+
  geom_sf(data=lkp, aes(color=depth_strata, fill= depth_strata), size=0.1)+
  geom_sf(data=esr%>%
            filter(Ecosystem_Area == "Gulf of Alaska"), 
          color="black", fill=NA)+
   scale_x_continuous(breaks = seq(0, 360, 10))+
  theme_bw()
```


And plot sample sizes for each region/depth

```{r, warning=FALSE, message=FALSE}
#plot sample sizes
lkp%>%
  group_by(ecosystem_sub, depth_strata)%>%
  summarize(n_sst=n())%>%
  ggplot()+
  geom_col(aes(x=depth_strata, y=n_sst, fill=depth_strata))+
  facet_wrap(~ecosystem_sub)+
  theme_bw()
```


## Plot subregional climatology by depth
First the GOA. This plot shows the average daily temperature for each region by depth. Solid lines are mean temps and dotted lines are standard deviations.

```{r, message = FALSE, warning=FALSE}
#load data from previous step
sst_10_200<-readRDS("data/sst_10_200.RDS")
sst_201_1000<-readRDS("data/sst_201_1000.RDS")
sst_1001<-readRDS("data/sst_1001.RDS")
sst<-sst_10_200%>%
  bind_rows(sst_201_1000)%>%
  bind_rows(sst_1001)

#summarize for plot
sst_sum<-sst %>%
  rename_with(tolower)%>%
  mutate(day=day(read_date),
         month=month(read_date),
         year=year(read_date),
         newdate=as.Date(paste("2000",month,day,sep="-"),format="%Y-%m-%d"))%>%
  group_by(ecosystem_sub, depth_strata, newdate)%>%
  summarise(meansst=mean(mean_sst),
            sd_sst=sd(mean_sst))


#plot goa
sst_sum%>%
  filter(ecosystem_sub%in% c("Eastern Gulf of Alaska", "Western Gulf of Alaska"))%>%
  ggplot()+
  geom_line(aes(x=newdate, y=meansst, color=depth_strata), size=1)+
    geom_line(aes(x=newdate, y=meansst+sd_sst, color=depth_strata), lty=2)+
    geom_line(aes(x=newdate, y=meansst-sd_sst, color=depth_strata), lty=2)+
  facet_wrap(~ecosystem_sub)+
        scale_x_date(date_breaks="1 month",
                   date_labels = "%b",
                   expand = c(0.025,0.025)) + 
  xlab("date")+
  ylab("CRWSST")
  theme_bw()
```

Next the Aleutians

```{r, include = FALSE}
#plot ai
sst_sum%>%
  filter(ecosystem_sub%in% c("Western Aleutians", "Central Aleutians", "Eastern Aleutians"))%>%
  ggplot()+
  geom_line(aes(x=newdate, y=meansst, color=depth_strata), size=1)+
    geom_line(aes(x=newdate, y=meansst+sd_sst, color=depth_strata), lty=2)+
    geom_line(aes(x=newdate, y=meansst-sd_sst, color=depth_strata), lty=2)+
  facet_wrap(~ecosystem_sub)+
        scale_x_date(date_breaks="1 month",
                   date_labels = "%b",
                   expand = c(0.025,0.025)) + 
  xlab("date")+
  theme_bw()

```
## Marine heatwave calculations
Here I plot presence and absence of heatwaves over time for each of the subregions and deoth 
```{r, message = FALSE, warning=FALSE}
#prepare sst data for heatwaveR code
sst<-sst %>% rename_with(tolower)   %>%
  filter(ecosystem_sub %in% c("Eastern Gulf of Alaska", "Western Gulf of Alaska"))%>%
         mutate(t=as.Date(read_date,format="%Y-%m-%d"),
                temp=mean_sst)

#function to detect mhw events for each region and depth strata
mhwfun<- function(x, region, strata) {
  (detect_event(ts2clm(x %>% filter(ecosystem_sub==region & depth_strata==strata)%>%
         arrange(t), climatologyPeriod = c("1985-01-01", "2014-12-31"))))$clim %>%
    mutate(ecosystem_sub=region, depth_strata=strata)
}

#test
egoa_10_mhw<-mhwfun(sst,"Eastern Gulf of Alaska", "-10 to -200")

#blank dataframe for loop
mhw_clim<-egoa_10_mhw[0,]

eco_vec<-unique(sst$ecosystem_sub)
depth_vec<-unique(sst$depth_strata)

#loop function
for (i in eco_vec) {
  for (j in depth_vec) {
mhw_clim<-mhw_clim%>%
  bind_rows(mhwfun(x=sst, region=i, strata=j))
  }
}

hcol<-"#ffc866"

#plotting function
mhwplotfun<-function(x, region){
  ggplot(data=x%>%filter(ecosystem_sub==region))+
  geom_rect(aes(xmin = t, xmax = t,
                  ymin = -Inf, ymax = Inf,
                  color = event)) +
  geom_line(aes(x=t, y=temp-seas))+
  facet_grid(rows=vars(depth_strata))+
  scale_color_manual(values=c("white", "#ffc866"))+
   #  scale_fill_manual(values=c("white", hcol))+
    ggtitle(paste0(region," heatwaves-CRWSST"))+
    scale_x_date(#limits=c(xmin, xmax), 
                 expand = c(0, 0))+
  theme(legend.position = "none",
        panel.border=element_rect(fill=NA))
}


ggplot(mhw_clim%>%filter(ecosystem_sub=="Eastern Gulf of Alaska"), aes(t, temp-seas)) + 
    geom_rect(aes(xmin = t, xmax = t,
                  ymin = -Inf, ymax = Inf,
                  color = event),
              size = 0.7,
              show.legend = FALSE) +
    geom_line()

#test
#mhwplotfun(mhw_clim, "Eastern Gulf of Alaska")

#plot all
for (i in eco_vec) {
  print(mhwplotfun(x=mhw_clim, region=i))
}

             
```

## COMPARISON WITH OISST

```{r, eval=FALSE, include=FALSE}
#extend timeout

#-10 to -200
oi_10_200<- dbFetch(dbSendQuery(con,
                           paste0("select b.ecosystem_sub, a.read_date, avg(a.temp) oi_sst from afsc.erddap_oi_sst a
left join afsc.erddap_oi_sst_spatial_lookup b on a.oi_id = b.id
where b.depth between -200 and -10
and b.ecosystem = 'Gulf of Alaska'
group by b.ecosystem_sub, a.read_date
order by b.ecosystem_sub, a.read_date")))

oi_10_200<-oi_10_200 %>% mutate(depth_strata="-10 to -200")

saveRDS(oi_10_200, "data/oi_10_200.RDS")

#-201 to -1000
oi_201_1000<- dbFetch(dbSendQuery(con,
                           paste0("select b.ecosystem_sub, a.read_date, avg(a.temp) oi_sst from afsc.erddap_oi_sst a
left join afsc.erddap_oi_sst_spatial_lookup b on a.oi_id = b.id
where b.depth between -1000 and -201
and b.ecosystem = 'Gulf of Alaska'
group by b.ecosystem_sub, a.read_date
order by b.ecosystem_sub, a.read_date")))

oi_201_1000<-oi_201_1000 %>% mutate(depth_strata="-201 to -1000")

saveRDS(oi_201_1000, "data/oi_201_1000.RDS")

# >1000
oi_1001<- dbFetch(dbSendQuery(con,
                           paste0("select b.ecosystem_sub, a.read_date, avg(a.temp) oi_sst from afsc.erddap_oi_sst a
left join afsc.erddap_oi_sst_spatial_lookup b on a.oi_id = b.id
where b.depth < -1000
and b.ecosystem = 'Gulf of Alaska'
group by b.ecosystem_sub, a.read_date")))

oi_1001<-oi_1001 %>% mutate(depth_strata="< -1000")

saveRDS(oi_1001, "data/oi_1001.RDS")

#spatial lookup table for plotting
dbFetch(dbSendQuery(con, paste0("select* from afsc.erddap_oi_sst_spatial_lookup"))) %>%
  saveRDS("data/oi_spatial_lookup.RDS")
```

```{r, message = FALSE, warning=FALSE}
#New version with just GOA
lkp<-readRDS("data/oi_spatial_lookup.RDS") %>%
  rename_with(tolower)%>%
  filter(depth <= -10 & ecosystem =="Gulf of Alaska") %>%
  mutate(depth_strata=ifelse(depth>=-200, "-10 to -200", ifelse(depth>= -1000, "-201 to -1000", "< -1000")),
         lon360=ifelse(longitude>0, longitude, longitude+360))%>%
  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326, agr = 'constant') %>%
  st_transform(crs=3338)

#map
ggplot()+
 # geom_sf(data=ak_dd%>%st_shift_longitude())+
  #geom_point(data=lkp, aes(x=lon360, y= latitude, color=depth_strata, fill= depth_strata))+
  geom_sf(data=lkp, aes(color=depth_strata, fill= depth_strata), size=1)+
  geom_sf(data=esr%>%
            filter(Ecosystem_Area == "Gulf of Alaska"), 
          color="black", fill=NA)+
   scale_x_continuous(breaks = seq(0, 360, 10))+
  ggtitle("OISST")+
  theme_bw()
```


```{r, warning=FALSE, message=FALSE}
#plot sample sizes
lkp%>%
  group_by(ecosystem_sub, depth_strata)%>%
  summarize(n_sst=n())%>%
  ggplot()+
  geom_col(aes(x=depth_strata, y=n_sst, fill=depth_strata))+
  facet_wrap(~ecosystem_sub)+
  ggtitle("OISST")+
  theme_bw()
```

```{r, message = FALSE, warning=FALSE}
#load data from previous step
sst_10_200<-readRDS("data/oi_10_200.RDS")
sst_201_1000<-readRDS("data/oi_201_1000.RDS")
sst_1001<-readRDS("data/oi_1001.RDS")
sst<-sst_10_200%>%
  bind_rows(sst_201_1000)%>%
  bind_rows(sst_1001)

#summarize for plot
sst_sum<-sst %>%
  rename_with(tolower)%>%
  mutate(day=day(read_date),
         month=month(read_date),
         year=year(read_date),
         newdate=as.Date(paste("2000",month,day,sep="-"),format="%Y-%m-%d"))%>%
  group_by(ecosystem_sub, depth_strata, newdate)%>%
  summarise(oisst=mean(oi_sst),
            sd_sst=sd(oi_sst))


#plot goa
sst_sum%>%
  filter(ecosystem_sub%in% c("Eastern Gulf of Alaska", "Western Gulf of Alaska"))%>%
  ggplot()+
  geom_line(aes(x=newdate, y=oisst, color=depth_strata), size=1)+
    geom_line(aes(x=newdate, y=oisst+sd_sst, color=depth_strata), lty=2)+
    geom_line(aes(x=newdate, y=oisst-sd_sst, color=depth_strata), lty=2)+
  facet_wrap(~ecosystem_sub)+
        scale_x_date(date_breaks="1 month",
                   date_labels = "%b",
                   expand = c(0.025,0.025)) + 
  xlab("date")+
  ylab("OISST")+
  theme_bw()
```

```{r, message = FALSE, warning=FALSE}
#prepare sst data for heatwaveR code
sst<-sst %>% rename_with(tolower)   %>%
         mutate(t=as.Date(read_date,format="%Y-%m-%d"),
                temp=oi_sst)

#function to detect mhw events for each region and depth strata
mhwfun<- function(x, region, strata) {
  (detect_event(ts2clm(x %>% filter(ecosystem_sub==region & depth_strata==strata)%>%
         arrange(t), climatologyPeriod = c("1985-01-01", "2014-12-31"))))$clim %>%
    mutate(ecosystem_sub=region, depth_strata=strata)
}

#test
egoa_10_mhw<-mhwfun(sst,"Eastern Gulf of Alaska", "-10 to -200")

#blank dataframe for loop
mhw_clim<-egoa_10_mhw[0,]

eco_vec<-unique(sst$ecosystem_sub)
depth_vec<-unique(sst$depth_strata)

#loop function
for (i in eco_vec) {
  for (j in depth_vec) {
mhw_clim<-mhw_clim%>%
  bind_rows(mhwfun(x=sst, region=i, strata=j))
  }
}

hcol<-"#ffc866"

#plotting function
mhwplotfun<-function(x, region){
  ggplot(data=x%>%filter(ecosystem_sub==region))+
  geom_rect(aes(xmin = t, xmax = t,
                  ymin = -Inf, ymax = Inf,
                  color = event)) +
  geom_line(aes(x=t, y=temp-seas))+
  facet_grid(rows=vars(depth_strata))+
  scale_color_manual(values=c("white", "#ffc866"))+
   #  scale_fill_manual(values=c("white", hcol))+
    ggtitle(paste0(region," heatwaves-OISST"))+
    scale_x_date(#limits=c(xmin, xmax), 
                 expand = c(0, 0))+
  theme(legend.position = "none",
        panel.border=element_rect(fill=NA))
}


ggplot(mhw_clim%>%filter(ecosystem_sub=="Eastern Gulf of Alaska"), aes(t, temp-seas)) + 
    geom_rect(aes(xmin = t, xmax = t,
                  ymin = -Inf, ymax = Inf,
                  color = event),
              size = 0.7,
              show.legend = FALSE) +
    geom_line()

#plot all
for (i in eco_vec) {
  print(mhwplotfun(x=mhw_clim, region=i))
}

             
```